Accelerating adaptive ultrasound algorithms using graphics processing units \vspace{0.5cm}

### 2014-02-05 ThesisJP/Introduction/introduction.tex
Small adjustments to the text.

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics Processing Unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision computations. For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

@@

The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging, an effective processing rate of around 100 MFLOPS is required. This number, as explained in Background \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.  Nevertheless, with modern GPUs, these levels of processing are finally available within a single card. When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

Main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}). However, two additional methods were also explored. The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}). Both share the property of being computationally complex, but consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.

@@

In this paper, to reduce the matrix size, it is taken advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace Capon beamforming}), which is implemented on the GPU. For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element Space Capon}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging. The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.  \textbf{Paper II} and \textbf{VI} also presents, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.

@@

\subsection{Paper V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we try to reduce the time for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version. Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.

>>>>

\section{Motivation}
A rapid development in computer game technology and accompanying programming languages have recently provided researchers with small personal super computers, comprised in a single graphics card (\nom{GPU}{Graphics processing unit}). The latest graphics cards\footnote{As of January 2014.} from both \nom{Nvidia}{Graphics card producer} and \nom{AMD}{Graphics card producer} provide close to 6 T\nom{FLOPS}{Floating point operations per second} single precision and 1.5 TFLOPS double precision in computational power. For double precision this is more than the worlds most powerful supercomputer could provide in 1998.

@@

The Capon beamformer  is a good example of an algorithm which has been highly studied for the last decade for the application of medical ultrasound imaging, but where the number of calculations was thought to be too high for real-time processing to happen any time soon. No wonder, to achieve real-time processing for e.g. cardiac ultrasound imaging, an effective processing rate of around 100 MFLOPS is required. This number, as explained in Section \ref{sec:adaptbf}, also grows with larger subarrays, an increasing number of samples, and higher frame rates.  Nevertheless, with modern GPUs, these levels of processing are finally available within a single card. When researchers are exploring new algorithms for ultrasound imaging it is therefore important to keep the architecture of parallel accelerators in mind. If a new complex algorithm is suppose to run in real time it needs to fit the programmable and parallel pipeline of modern ultrasound scanners.

@@

Main focus have been the Capon beamformer, and the problem of making this computationally intense algorithm available for real-time ultrasound imaging (\textbf{Paper\,I} and \textbf{II}). However, two additional methods were also explored. The first is adaptive visualization of cardiac ultrasound volumes (\textbf{Paper\,IV}) and the latter is simulation of dense ultrasound pressure fields (\textbf{Paper\,V}). Both share the property of being computationally complex, but on the other hand they consist of many independent computations which makes them perfectly suited for parallel GPU acceleration.

@@

In this paper, to reduce the matrix size, it is taken advantage of the beamspace version of the Capon beamformer (\nom{BS-Capon}{Beamspace capon beamforming}), which is implemented on the GPU. For parameters previously derived to give similar performance for BS-Capon and element-space Capon (\nom{ES-Capon}{Element space capon beamforming}), we show that real-time BS-Capon beamforming is feasible for cardiac ultrasound imaging. The reported processing throughput is able to keep up with the acquisition frame rate in a typical cardiac ultrasound imaging system equipped with 64 and 96 element linear arrays.  \textbf{Paper II} and \textbf{VI} also presents, for the first time, videos where the ES-Capon and BS-Capon beamformer have been applied on loops of simulated and \textit{in vivo} medical ultrasound images.

@@

\subsection{Paper V}
\textbf{Huygens on Speed: Interactive Simulation of Ultrasound Pressure Fields}\\
\textbf{J.\:P.\:\AA{}sen} and S.\:Holm\\
{\it Proc. IEEE Ultrasonics Symposium 2012 \todo{add pages}.}\\\\
The fifth paper presents an interactive simulations tool capable of simulating dense pressure fields from ultrasound arrays in real time. Simulation tools are heavily used by researchers in order to test out new algorithms and to deduce the performance of new array designs prior to manufacturing. However, dense simulations involve extensive calculation and computational time. With the presented simulator we therefore try to reduce the time for calculating pressure fields by means of GPU acceleration. The tool is based on Huygens' principle which describes diffraction caused by a slit as the superposition of several point sources located inside the opening. Thus, the simulator works by accumulating the contribution from a collection of point sources in a set of observation points. How the source and observation points are laid out is left for the user to decide. Point sources positioned along a line will for instance simulate an ultrasound array. The simulator is linked with both a Paint-like interface for interactive drawing of arrays, and a Matlab interface for precise scripting of array configurations. The main contributions in this paper is the increased performance of the GPU implementation compared with a CPU and Matlab\footnote{The Ultrasim toolbox \todo{add citation}} version. Second, the paint-like interface provides a neat way of demonstrating array beamforming principle in real time. We believe this to be of great value both for array designers and teachers.
